Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-05-24 docs/parasite_mortality_manuscript.tex
Adding in Cherie's comments

Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
induced mortality, particularly for macroparasites where pathology is
linked to to the intensity of infection, can provide important insight into
parasite transmission and host population management. This is, however,
notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
quantifying parasite-induced mortality in wildlife populations from intensity
data and, using simulations, we show that our method is more reliable for
detecting and quantifying parasite-induced host mortality than the methods that are
currently available.  We also show that this method provides quantitative
assessments of parasite-induced mortality in empirical data that are consist
with previously published qualitative assessments. However, we stress that this
method, along with all methods for estimating parasite-induced mortality in
host populations, have a number critical assumptions that limit their
applicability in the real world.  We conclude that methods for estimating and
detecting parasite-induced mortality from intensity data should only be used as an
exploratory tool for informing more rigorous studies of parasite-induced host
mortality.

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
This method is sufficient for many bacterial and viral agents that reproduce within a
host, however for macroparasites pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013},
wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced mortality. We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in these heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data.

The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
probability. \cite{Adjei1986} suggested that this relationship
could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
probability of host survival with increasing parasite intensity. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
This method appeared to provide an estimate for the parasite intensity at which
50\% of hosts exhibit PIHM ($LD_{50}$), as well as the unmeasurable fraction of the
population that was lost (\emph{SI} 2). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small.

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule, it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function given by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and has never been tested.

Intensity data should only
be used to estimate parasite impacts on host populations if unbiased and accurate methods exist. In
this study, we first propose a novel method for detecting and quantifying PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, and  uses standard statistical techniques to determine PIHM significance. We provide Python code for implementing the Likelihood Method in \emph{SI} 4.

As with all previously proposed methods for estimating PIHM, the
Likelihood Method first assumes that prior to mortality the parasite distribution is described by a negative binomial $g(x; \mu_p, k_p)$, where $\mu_p$ and $k_p$ are the mean parasite intensity and aggregation before mortality, respectively (smaller $k_p$ indicates more aggregation).   Previous methods required calculating the total number of hosts before mortality ($N_p$) \citep{Crofton1971a,Adjei1986}, however this parameter is not needed in the Likelihood Method.

The Likelihood Method then assumes that the host survival function, which specifies the probability of a host surviving with $x$ parasites, follows the logistic curve given by

\begin{equation}
    h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
    \label{eq:logistic}
\end{equation}
With these two explicit assumptions, the Likelihood Method estimates four parameters: $\mu_p$, $k_p$, $a$, and $b$ by  first defining a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.  Using standard rules of conditional probability this distribution can be written as

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(x; a, b)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \mu_p, k_p)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(x; a, b)  * g(x; \mu_p, k_p)}{\sum_{x=0}^{\infty} h(x; a, b)  * g(x; \mu_p, k_p)}
    \label{eq:dist}
\end{equation}

Using this probability distribution, one can then find the parameters $\mu_p$, $k_p$, $a$, and $b$ that maximize the likelihood of an observed host-parasite dataset. The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.  All parameters are defined in Table \ref{tab:params}.

To estimate the significance of PIHM in a host-parasite system, a
likelihood ratio test is used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by a negative binomial
distribution.  If PIHM is not significant in the system, the resulting likelihood
ratio statistic should approximately follow a $\chi^2$ distribution with two degrees of freedom.

>>>>

Parasites can often have significant impacts on host populations through parasite-induced host mortality.  Detecting and quantifying the magnitude of parasite-
induced mortality, particularly for macroparasites in which pathology is
linked to to the intensity of infection, can provide important insight into
parasite dynamics and host population management. This is, however,
notoriously difficult in practice as the only data often available are measures of parasite intensities. In this study we provide a novel method for detecting and
quantifying parasite-induced mortality in wildlife populations from intensity
data and, using simulations, we show that our method is more reliable for
detecting and quantifying parasite-induced host mortality than the methods that are
currently available.  We also show that this method provides quantitative
estimates of parasite-induced mortality in empirical data that are consist
with previously published qualitative estimates. However, we stress that this
method, along with all methods for estimating parasite-induced mortality in
host populations for intensity data alone, has a number critical assumptions that limit their
applicability in the real world.  We conclude that methods for estimating and
detecting parasite-induced mortality from intensity data should be used only as an
exploratory tool for informing more rigorous studies of parasite-induced host
mortality.

@@

Infectious agents can have major impacts on animal populations through changing
population dynamics and stability \citep{Dobson1992}, altering predator-prey interactions \citep{Joly2004}, and
even causing species' decline and extinction \citep{DeCastro2005a,McCallum2012b}. Accurately estimating the impact
of these infectious agents in wildlife is critical to understanding what
regulates host and parasite populations, making predictions about disease
transmission, and managing disease outbreaks \citep{Langwig2015}. The impact of microparasite pathogens, such as rabies \citep{Coyne1989}, bovine TB \citep{Cox2005}, and
rinderpest \citep{Tille1991}, is typically quantified based on the presence or absence of
disease, and does not account for the number of infectious agents present.
This method is sufficient for many bacterial and viral agents that reproduce within a
host, however for macroparasites, pathology is linked to the intensity of infection and hosts cannot be simply categorized as infected and
uninfected \citep{AndersonandMay1979,Lafferty2002}.  Helminths exhibiting this intensity dependent pathology have
significant impacts on human health \citep{Brooker2004}, domestic livestock economics \citep{Roeber2013}, and
wildlife survival \citep{Kirk2003, Logiudice2003}.  While it is generally assumed that some fraction of
wild host populations must succumb to parasitic infections, it is notoriously
difficult to actually quantify parasite-induced host mortality (PIHM) in wild
animal populations \citep{McCallum2000a}.

Ideally, parasite-induced host mortality is
quantified by experimentally infecting and tracking individual hosts in the
wild population; however, for logistical and ethical reasons this method is
rarely feasible \citep{McCallum2000a}. Data on parasite intensity is much easier to collect and has
often been used to identify the presence of PIHM \citep{Crofton1971a,Lester1977,Lester1984,Lanciani1989,Royce1990,Ferguson2011} and to quantify the
relationship between infection intensity and host mortality \citep{Adjei1986}.

\cite{Crofton1971a} first proposed that PIHM could be identified by comparing the
observed parasite distribution in the host population to the distribution
predicted in the absence of parasite-induced mortality. We briefly introduce the Crofton Method here and provide a more detailed explanation of its implementation in \emph{Supplementary Material (SI)} 1. This method
assumes that, prior to host mortality, infection intensity in the host population follows a negative binomial distribution and the tail of the distribution is truncated as intensity dependent pathology removes the most heavily infected hosts. Assuming mortality occurs only in heavily infected hosts, evidence of this parasite-induced mortality should then be detectable by iteratively
fitting a negative binomial distribution to hosts with lower and lower parasite
loads, and comparing these truncated predicted distributions to the corresponding truncated observed parasite data. [FIGURE]

The Crofton Method may be able to detect the presence of PIHM, but it does not quantify the relationship between infection intensity and host survival
probability. \cite{Adjei1986} suggested that this relationship
could be calculated by first using the Crofton Method to estimate the pre-mortality parasite distribution and then using this distribution to calculate the
probability of host survival with increasing parasite intensity. To do this,
\cite{Adjei1986} modeled host survival as a logistic function and then used a generalized linear model (GLM) to estimate the logistic parameters (see \emph{SI} 2 for a technical description of the Adjei Method).
\citeauthor{Adjei1986} suggested that this method could provide an estimate for the parasite intensity at which a host has a 50\% chance of suffering parasite-induced mortalith ($LD_{50}$). However, to implement this method the observed data must be
modified to fit the GLM framework and subjectively binned when mean
infection intensity is high or sample sizes are small (see SI 2 for details).

After 30 years, and despite clear limitations \citep{McCallum2000a}, these
methods (particularly the Crofton Method) are still discussed among
parasitologists and are the primary techniques for examining population level
impacts of parasitism using parasite intensity data. In these methods, PIHM can
only be identified by visually examining plots and, with no clear decision
rule, it can be difficult to determine the significance of PIHM across
different host-parasite systems. The survival function given by the Adjei
Method offers one solution; however, this method requires manipulating the
original data and its accuracy has never been validated.

Intensity data should
be used to estimate parasite impacts on host populations only if unbiased and accurate methods exist. In
this study, we first propose a novel method for detecting and quantifying PIHM. We
next use simulations to compare our method with the previous Adjei
Method to test the ability of both to (1) detect occurrence of PIHM and (2) estimate the
lethal parasite load ($LD_{50}$) and the associated survival function.  We then
apply both methods to real datasets previously used in PIHM analyses and
compare the results. Finally, we discuss the limitations of inferring PIHM
from intensity data and whether any method for inferring PIHM has a place
in quantitative parasitology.

@@

Here we propose a novel, likelihood-based method (henceforth Likelihood Method) that does not
require binning or data alteration, reduces the number of
parameters to be estimated, is highly generalizable, and uses standard statistical techniques to determine PIHM significance.
%We provide Python code for implementing the Likelihood Method in \emph{SI} 4.
The Likelihood Method first assumes that prior to mortality the parasite distribution can be described by the distribution $g(x; \phi)$, which specifies the probability of a host having $x$ parasites when it is observed.  $\phi$ is a vector of parameters that described the shape and scale of this distribution.

The Likelihood Method then assumes that the probability of a host surviving for
$t$ units given it has some parasite intensity $x$ is described by the function
$h(\text{survival} ; x, t, \theta)$ where $\theta$ is a vector of parameters of
the survival function. For most parasite-intensity datasets, the observer has
no knowledge of how long a host has been infected with $x$.  One way to account
for this would be to integrate out time such that we are left with the function
$h(\text{survival} ; x, \theta)$ specifying the probability of survival given
$x$ parasites and some additional parameters $\theta$.  While eliminating the
functions dependence on time is far from ideal, this is an inherent limitation
of attempting inference on cross-sectional data (jargon).

With these two assumptions, we can define a probability distribution that gives the probability of having a parasite load of $x$ parasites conditional on host survival.  Using standard rules of conditional probability this distribution can be written as

@@

One can see that $P(\text{survival} | x)$ is the survival function $h(\text{survival}; x, \theta)$, $P(x)$ is the pre-mortality parasite distribution $g(x; \phi)$ and $P(\text{survival}) = \sum_{x=0}^{\infty} P(\text{survival} | x) * P(x) =  \sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)$. Therefore equation \ref{eq:concept} can be written as

\begin{equation}
    P(x | \text{survival}) = \dfrac{h(\text{survival}; x, \theta)  * g(x; \phi)}{\sum_{x=0}^{\infty} h(\text{survival}; x, \theta)  * g(x; \phi)}
    \label{eq:dist}
\end{equation}

% \begin{equation}
%     h(x ; a, b) = \dfrac{e^{a - b \log(x)}}{1 + e^{a - b \log(x)}}
%     \label{eq:logistic}
% \end{equation}

Using this probability distribution, one can then find the parameters $\theta$
and $\phi$ that maximize the likelihood of an observed host-parasite dataset.
To estimate the significance of PIHM in a host-parasite system, a likelihood
ratio test can be used in which the full model is given by equation
\ref{eq:dist} and the reduced model is given by the pre-mortality distribution
$g(x; \phi)$.  If PIHM is not significant in the system, the resulting
likelihood ratio statistic should approximately follow a $\chi^2$ distribution
with two degrees of freedom.

Equation \ref{eq:dist} could be parameterized in many different ways depending
on the parasite system of interest. In this study, we will follow the precedent
set by all previous methods for estimating PIHM and assume that the pre-
mortality parasite distribution $g(x; \phi)$ follows a negative binomial
distribution with the parameters mean parasite intensity ($\mu_p$) and
aggregation ($k_p$) before mortality, respectively (smaller $k_p$ indicates
more aggregation).   The negative binomial distribution is not just a
phenomenological assumption and can arise as the equilibrium parasite
distribution under a variety of different biological assumptions
\citep{Calabrese2011} [MORE]. However, it is also an incredibly flexible
distribution that fits many host-parasite systems regardless of whether the
underlying mechanisms lead to an exact negative binomial distribution
\citep{Shaw1998}.

Choosing a function for $h(\text{survival}; x, \theta)$

 The equation $\exp(a / b)$ can then be used to calculate the parasite $LD_{50}$, here defined as the infection intensity at which 50\% of hosts experience PIHM.  All parameters are defined in Table \ref{tab:params}.
